A Machine Learning Model Built with Time-Dependent Lab Data to Predict Chronic Urticaria Duration and Provide Clinical Insights

### 2024-10-25 latex/manuscripts.tex
241025

%% 
%% Copyright 2007-2024 Elsevier Ltd
%% 
%% This file is part of the 'Elsarticle Bundle'.
%% ---------------------------------------------
%% 
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.3 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.3 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%% 
%% The list of all files belonging to the 'Elsarticle Bundle' is
%% given in the file `manifest.txt'.
%% 
%% Template article for Elsevier's document class `elsarticle'
%% with harvard style bibliographic references

@@

\title{Machine learning model for predicting the visit duration of urticaria based on clinical laboratory data}

@@

%% Abstract
\begin{abstract}
%% Text of abstract
Background: Urticaria is a common condition presenting with wheals, angioedema, or both, driven by mast cell degranulation. The duration of urticaria, particularly chronic spontaneous urticaria (CSU), is crucial for effective patient management and treatment planning. This study aimed to build a machine learning model for predicting the visit duration of urticaria based on clinical laboratory data and to identify the factors that contribute to prolonged episodes of chronic urticaria.

Method:

Results:

Conclusion: 
\end{abstract}

@@

Chronic urticaria significantly impairs quality of life, affecting work and school performance. It is considered a severe allergic disease due to its disabling nature and high disease burden\citep{Zuberbier2021The}. Duration of CSU greater than 3 years are associated with better responses to second-generation antihistamines and other treatments\citep{Chiang2022Predictors}. Predicting the duration of urticaria, particularly chronic spontaneous urticaria (CSU), is crucial for effective patient management and treatment planning.

@@

\section{Methods}\label{Methods}
%% Use \subsection commands to start a subsection.
\subsection{Patients}\label{Patients}

Patients with urticaria were recruited from the urticaria specialty clinic of the dermatology department of Shanghai XinHua Hospital affiliated to Shanghai Jiao Tong University School of Medicine from January 2018 to December 2024. The inclusion criteria were as follows: (1) patients diagnosed with urticaria according to the EAACI/GA2LEN/EDF/WAO guidelines\citep{Zuberbier2021The}; (2) patients with complete clinical and laboratory data; (3) patients with stable follow-up history, indicated by at least 3 times of follow-up visits. the exclusion criteria were as follows: (1) patients with other skin diseases; (2) patients with severe systemic diseases; (3) patients with incomplete clinical data. The study was approved by the ethics committee of Shanghai XinHua Hospital affiliated to Shanghai Jiao Tong University School of Medicine, and all patients provided written informed consent.

@@

Data of patients with urticaria were collected from the electronic medical record system of the hospital, including demographic data, clinical data, laboratory data. 
the data were stored in a Mysql database for subsequent analysis, with schema as follows: (1) Patients: containing basic information of each unique patient; (2) OutpatientNumbers: storing relationship between outpatient numbers with unique patient; (3) PatientVisits: containing visit events records; (4) PatientExaminations: containing the examination events records; (5) ExaminationItems: a dictionary table describing the examination items. the database schema is shown in database. Database markup language (DBML) describing the database schema is shown in supplementary materials \ref{file:DBschema}

\subsection{Feature extraction and feature engineering}\label{FeatureEngineering}
Visit duration, calculated by the difference between the first visit date and the last visit date in the urticaria specialty clinic, was the target variable. We suppose that the visit duration of a patient at the urticaria specialty clinic is a surrogate for the disease duration of urticaria.
The following features were extracted from the database: (1) demographic data: gender, first visit age; (2) clinical data: whether the patient having concomitant inducible urticaria; (3) laboratory data: results from common blood tests, CRP, immunoglobulin, 25-hydroxyvitamin D, Thyroid function, autoantibodies, coagulation function, common urine tests, and allergen specific IgE tests. 
For laboratory data, 2 types of data were extracted: time-independent data and time-dependent data. Time-independent data are average values of laboratory data during the whole follow-up period, while time-dependent data are average values of laboratory data durting preclincial phase (before the onset of urticaria), acute phase (within 6 weeks after the onset of urticaria), and chronic phase (after 6 weeks of the onset of urticaria). 2 datasets were generated: one with time-independent data and one with time-dependent data, and were compared for prediction performance in the model development process. The sql queries for data are shown in in supplementary materials \ref{file:sql} and \ref{file:sql_time}.

\subsection{Model development and comparison}\label{Training}
Dataset was split into training set and external test set with a ratio of 7:3. 5 models were adopted for comparison: Xgboost, random forest, gradient boosting machine (GBM), adaboost, and support vector machine (SVM). 
Time-dependent data extraction may introduce multicollinearity between features, which can affect model performance. To address this issue, L1 regularization, L2 regularization, and column sampling methods were employed in XGBoost to handle multicollinearity. L1 regularization helps eliminate irrelevant collinear features by sparse feature selection, while L2 regularization reduces the model's dependence on collinear features by smoothing the weights. Column sampling reduces the likelihood of collinear features being selected simultaneously by randomly choosing a subset of features when building each tree, which is helpful when interpreting the model using method like SHAP, since it enables more balanced weight distribution across features and reduces the possibility of one feature overshadowing others. The Random Forest model does not support L1 and L2 regularization, but it handles multicollinearity through the max\_features parameter, which is similar to column sampling. L2 regularization was applied to the SVM model as well. Additionally, the frameworks of XGBoost and GBM incorporate greedy split and boosting mechanisms, selecting the most important features for each tree and fitting the residuals step by step, allowing the model to focus on the key features rather than over-relying on collinear ones. 
Hyperparameter optimization on 10 model settings (5 models and 2 datasets) was performed using TPE algorithm by nni package in python, which is a bayesian optimization algorithm that uses tree-structured parzen estimator to model the objective function and suggest the next set of hyperparameters to evaluate based on the previous results. Internal 5 fold cross-validation was employed to discern the most suitable hyperparameters for each distinct model, individually applied to each model settings for enhanced performance. The performance of the model and data was evaluated by receiver operating characteristic (ROC) curve, area under the curve (AUC), accuracy, precision, recall and F1 score on different cuttoffs of disease duration. The sensitivity, specificity, PPV, NPV, accuracy, and F1 score were calculated at the optimal cutoff value that maximized the Youden index. The model and data with the best performance were selected for further analysis.

@@

In this study, Boruta algorithm was performed on the final model, with max iteration set to 50 and repeatition for 15 times. The confirmed features list and importance rank were recorded for each run. The overlap of confirmed features across runs was calculated to determine the final feature list. For time dependent data, once a laboratory item was confirmed as important in any phase of disease, the feature of the same lab item in other phase of disease was also considered as important. That is to say, when top 10 items were selected, the features representing different phases of that item were all included in the final model, which is at most 30 features for time dependent data. The model performance on different number of top items, ranked by the average importance score, were compared to determine the optimal number of features for the final model.

\subsection{Model explanation}\label{ModelExplanationMethods}
SHAP (Shapley Additive Explanations) is an explanation method based on game theory, used to interpret the outputs of machine learning models. It assigns each feature's contribution to the model's prediction, helping users understand the decision-making process of the model. In tree models that are not sensative to collinearity, such as Xgboost and GBM, the model tend to select one most important feature among collinear features for a split and ignores the other correlated features. As a result, SHAP may assign a high contribution to the first selected feature while the others receive little or no contribution. This happens because the model captures most of the information through the first feature, effectively overshadowing the others. By introducing joint distribution estimation for correlated features during SHAP calculations, we can capture their dependencies and ensure that SHAP values reflect the true impact of correlated features \citep{lundberg2017unified}.
The shapley value of each laboratory item in different phase of disease were compared to reveal various predicting ability. 
Some of the features were further analyzed with different age populations to explore the potential age effect on the predicting ability of laboratory data.

\section{Results}\label{Results}

@@

Of the 9921 patients who visited the urticaria specialty clinic and met the inclusion criteria from January 2018 to December 2024. 3954 patients were selected for time independent dataset and 1948 patients were selected for time dependent dataset, the rest were excluded due to serious missing data. 3954 patients in time independent data and 1948 patients time dependent data were allocated into seperate derivation and external validation sets with a ratio of 7:3, respectively. The comparison of demographic and clinical variables among the training and external validation sets is shown in Table \ref{tab:train_test_origi} for time independent dataset and Table \ref{tab:train_test_time} for time dependent dataset. No substantial differences were observed between the training set and the external test set across the majority of features in either time independent or time dependent data. Details of the study design are displayed in Figure \ref{protocol}.

% insert image
\begin{figure}[t]
    \centering
    \includegraphics[width=\textwidth]{figures/protocol.png}
    \caption{Study design}\label{protocol}
\end{figure}

Table \ref{tab:good_outcome_poor_outcome_origi} delineates the disparities between groups regarding the disease duration of patients in time independent Dataset. Among the 3954 patients, 2027 patients had a visit duration of less than 100 days, while 1927 patients had a visit duration of 100 days or more. The mean visit duration was 25.37 days for the short duration group and 604.32 days for the long duration group. The comparison of clinical characteristics between patients with short and prolonged visit duration in the time-independent dataset (Table \ref{tab:good_outcome_poor_outcome_origi}) highlights several trends that align with findings from previous studies on chronic spontaneous urticaria (CSU). In this analysis, immunoglobulin E (IgE) ($131.65 \pm 242.00$ in good outcome vs $152.28 \pm 346.75$ in poor outcome), eosinophils percentage ($1.55 \pm 2.02$ in good outcome vs $2.31 \pm 2.37$ in poor outcome), eosinophils counts ($128.60 \pm 141.68$ in good outcome vs $162.83 \pm 190.02$ in poor outcome), and basophil percentage ($0.25 \pm 0.22$ in good outcome vs $0.32 \pm 0.24$ in poor outcome) were significantly higher in patients with poor outcomes, which is consistent with previous studies that have linked these markers to CSU severity and duration \citep{SanchezBorges2017Factors,Rabelo-Filardi2013Parameters, Kolkhir2019Eosinopenia}. In contrast, C-reactive protein (CRP) levels were significantly lower in patients with poor outcomes ($13.03 \pm 19.62$ in good outcome vs $7.71 \pm 10.23$ in poor outcome), which contradicts previous findings that have associated higher CRP levels with CSU severity \citep{Rabelo-Filardi2013Parameters}. However, this discrepancy could reflect the complexity of CRP as an inflammatory marker that may vary across different disease stages or subgroups of patients. In time dependent data, which is shown in supplementary Table \ref{tab:good_outcome_poor_outcome_time}, CRP presented various trends in different phases of disease, with significantly lower CRP levels in acute phase ($10.21 \pm 10.95$ in good outcome vs $8.95 \pm 8.22$ in poor outcome) and higher CRP levels in chronic phase ($5.39 \pm 4.84$ in good outcome vs $6.47 \pm 8.81$ in poor outcome) in patients with poor outcomes. This finding suggests that the predictive ability of CRP may be influenced by the disease phase, which could explain the inconsistent results observed in the time-independent dataset.

@@

    \makecell[l]{CR eactive Protein} & $13.03 \pm 19.62$ & $7.71 \pm 10.23$ & 0.000 $\downarrow$ \\

@@

    \makecell[l]{Ro 52} & $2.10 \pm 5.90$ & $2.23 \pm 6.81$ & 0.515  \\
    \hline\end{tabular}\caption{Comparison of the characteristics between patients with good and poor outcomes in the time independent dataset \\ continuous variables are presented as mean ± standard deviation, categorical variables are presented as number (percentage) \\ good outcome is defined as visit duration < 100 days, poor outcome is defined as visit duration $\geq$ 100 days} \label{tab:good_outcome_poor_outcome_origi}
    \end{table}

\subsection{Comparison of multiple models on time-dependent and time-independent data}\label{ModelComparison}
% xgboost + time dependent datasset is the best model in 42 100 365 cutoffs

Time independent and Time dependent data, combined with 5 different models (XGBoost, Random Forest, AdaBoost, GBM, SVM), were used to generate 10 machine learning models to predict the disease duration of urticaria. Although most models utilized in this study is based on regression trees, their performance in directly predicting continuous numeric values in the regression task was suboptimal, which is likely due to the complexity of the underlying data distribution and model constraints. The predicted values, while not accurate for continuous outcome prediction, can still be utilized for binary classification tasks by applying an optimal threshold. Specifically, the model’s output is better suited for classification of disease duration when applying a predefined threshold that maximizes the classification performance (e.g., based on ROC-AUC or Youden's index). Therefore we use binary classification task to evaluate the model performance. Among the 10 models, the XGBoost model with time-dependent data (AUC = $0.906 \pm 0.013$ for 42, AUC = $0.926 \pm 0.013$ for 100, AUC = $0.884 \pm 0.018$ for 365) has the best predictive effect for disease duration, followed by the Random Forest model with time-dependent data (AUC = $0.933 \pm 0.025$ for 42, AUC = $0.922 \pm 0.028$ for 100, AUC = $0.837 \pm 0.037$ for 365). The ROC curves for the top three best-performing ML models are presented in Figure \ref{kfoldfirst3}. The discriminative performances and ROC curves of all 10 models are listed in Supplementary Table \ref{tab:kfold_results}, and Supplementary Figure \ref{fig:kfold_all}, respectively.

% insert image
\begin{figure}[t] 
    \centering 
    \includegraphics[width=\textwidth]{figures/kfoldfirst3.png} 
    \caption{ROC curves for 5-fold cross-validation results for three different models (XGBoost, Random Forest, and GBM) at three binary classification thresholds (42, 100, and 365). Each row (A, B, and C) represents the results for a specific threshold. The models' performances are plotted with time-independent and time-dependent datasets, where red indicates the time-independent dataset and blue represents the time-dependent dataset. The shaded areas represent the 95\% confidence intervals.}\label{kfoldfirst3} 
\end{figure}

@@

Xgboost with time dependent data, as the best performing model, was selected for further optimization. Feature importance ranking by Boruta algorithm are shown in Figure \ref{boruta_by_group} and Supplementary Table \ref{tab:boruta_ranking_df}. A total of 38 laboratory items were confirmed as important by Boruta algorithm, as shown in Supplementary Table \ref{tab:boruta_confirmed_vars}. During the process of reducing features based on the feature importance, the changes in AUCs for the model shows that the model maintain a robust performance unitil the number of items reduced to less than 10, as shown in Figure \ref{topn_extval25} A-C. The 25-item model shows the best performance in internal validation set with AUC of 0.914, 0.939, 0.899 for 42, 100, 365 cutoffs, respectively. The 25-item model were selected as the final model, and showed a stable performance in external test set with AUC of 0.887, 0.898, 0.809 for 42, 100, 365 cutoffs, as shown in Figure 3 \ref{topn_extval25} D-F.

% insert image
\begin{figure}[t] 
    \centering
    \includegraphics[width=\textwidth]{figures/boruta_by_group.png} 
    \caption{Boxplot representing averaged feature ranking distribution for each laboratory item generated by the Boruta algorithm over 15 iterations. The ranking value for each laboratory item is averaged for 3 different phases of disease. The y-axis shows the importance rank, with lower values indicating higher importance, while the x-axis lists the laboratory items names.}\label{boruta_by_group}
\end{figure}

% insert image
\begin{figure}[t]  
    \centering 
    \includegraphics[width=\textwidth]{figures/topn_extval25.png} 
    \caption{Panels A, B, and C display the ROC curves for the model as the number of top items is reduced based on the average importance score at cutoff values of 42, 100, and 365, respectively. Panels D, E, and F show the ROC curves for the final model with 25 top items on external test set at cutoff values of 42, 100, and 365, respectively.}\label{topn_extval25}

@@

\subsection{Model explanation}\label{ModelExplanationResults}
The SHAP analysis was performed on the final model with 25 top items. The heatmap shows shap values of the top 25 features on 300 samples from dataset, with the samples ordered by hierarchical clustering based on the similarity of shap values, and the corresponding model output and global importance of each feature, shown in Figure \ref{heatmap_all_orderx}.

\begin{figure}[t] 
    \centering
    \includegraphics[width=\textwidth]{figures/heatmap_all_orderx.png}
    \caption{a heatmap plot with the instances on the x-axis, the model inputs on the y-axis, and the SHAP values encoded on a color scale. The samples are ordered based on a hierarchical clustering by their explanation similarity. This results in samples that have the same model output for the same reason getting grouped together. The output of the model is shown above the heatmap matrix, and the global importance of each model input shown as a bar plot on the right hand side of the plot.}\label{heatmap_all_orderx}
\end{figure}

@@

% insert image
\begin{figure}[t] 
    \centering
    \includegraphics[width=\textwidth]{figures/modelexplanation.png} 
    \caption{}\label{modelexplanation}
\end{figure}

@@

\href{run:data.sql}{Supplementary file 2: SQL queries for time independent extraction}
\label{file:sql}

\href{run:data_time.sql}{Supplementary file 3: SQL queries for time dependent extraction}
\label{file:sql_time}

>>>>

%% Copyright 2007-2024 Elsevier Ltd
%% 
%% This file is part of the 'Elsarticle Bundle'.
%% ---------------------------------------------
%% 
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.3 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.3 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%% 
%% The list of all files belonging to the 'Elsarticle Bundle' is
%% given in the file `manifest.txt'.
%% 
%% Template article for Elsevier's document class `elsarticle'
%% with harvard style bibliographic references

@@

\title{A Machine Learning Model Built with Time-Dependent Lab Data to Predict Chronic Urticaria Duration and Provide Clinical Insights}

@@

%% Abstract
\begin{abstract}
%% Text of abstract
Background: Chronic spontaneous urticaria (CSU) is a persistent condition characterized by wheals and/or angioedema that significantly impacts patients' quality of life. Predicting the duration of CSU is crucial for effective management and treatment planning. While several biomarkers have been linked to CSU severity and duration, there is limited research on integrating clinical and laboratory data to build predictive models for disease prognosis.

Objective: This study aims to develop a machine learning model to predict visit duration in patients with urticaria based on clinical and laboratory data, and to identify key factors contributing to prolonged episodes of chronic urticaria.

Methods: A retrospective cohort of patients with urticaria from a dermatology clinic was analyzed. Time-independent and time-dependent laboratory data were extracted from electronic medical records. Five machine learning models (XGBoost, Random Forest, Gradient Boosting Machine, AdaBoost, and Support Vector Machine) were trained to predict visit duration, with binary classification thresholds set at 100 and 365 days. Model performance was evaluated using ROC-AUC, accuracy, precision, recall, and F1 score. SHAP analysis was conducted to interpret the contribution of individual features to model predictions, and the Boruta algorithm was used for feature selection.

Results: The XGBoost model with time-dependent data outperformed other models, achieving an AUC of 0.932 for the 100-day threshold. SHAP analysis identified basophils, eosinophils, C-reactive protein (CRP), and immunoglobulin E (IgE) as key predictors of prolonged visit duration, with phase-specific trends in their predictive power. Age-dependent effects were also observed, with certain laboratory markers showing differential impacts in younger versus older patients.

Conclusion: Our machine learning model successfully predicts visit duration in urticaria patients, providing insights into key clinical and laboratory markers that influence disease prognosis. The findings offer a valuable reference for personalized treatment strategies and highlight the importance of phase- and age-specific approaches in managing chronic urticaria.
\end{abstract}

@@

Chronic urticaria is considered a severe allergic disease with high disease burden\citep{Zuberbier2021The}. Epidemiologic studies have established that half of the patients with chronic urticaria achieved remission after 11 months, and approximately 90\% achieved remission within 5 years\citep{eun2019natural}. Several studies have indicated that the duration and disease severity of CSU are associated. It is observed that longer disease duration and positive ASST were present in patients with more severe symptoms\citep{SanchezBorges2017Factors}. Predicting the duration of chronic urticaria, is crucial for effective patient management and treatment planning.

@@

\section{Methods}\label{Methods}

\subsection{Patients}\label{Patients}

Patients with urticaria were recruited from the urticaria specialty clinic of the dermatology department of Shanghai XinHua Hospital affiliated to Shanghai Jiao Tong University School of Medicine from January 2019 to December 2023. The inclusion criteria were as follows: (1) patients diagnosed with urticaria according to the EAACI/GA2LEN/EDF/WAO guidelines\citep{Zuberbier2021The}; (2) patients with complete clinical and laboratory data; (3) patients with stable follow-up history, indicated by at least 3 times of follow-up visits. the exclusion criteria were as follows: (1) patients with other skin diseases; (2) patients with severe systemic diseases; (3) patients with incomplete clinical data. The study was approved by the ethics committee of Shanghai XinHua Hospital affiliated to Shanghai Jiao Tong University School of Medicine, and all patients provided written informed consent.

@@

Data of patients with urticaria were collected from the electronic medical record system of the hospital, including demographic data, clinical data, laboratory data. The data were stored in a Mysql database for subsequent analysis, with schema as follows: (1) Patients: containing basic information of each unique patient; (2) OutpatientNumbers: storing relationship between outpatient numbers with unique patient; (3) PatientVisits: containing visit events records; (4) PatientExaminations: containing the examination events records; (5) ExaminationItems: a dictionary table describing the examination items. the database schema is shown in database. Database markup language (DBML) describing the database schema is shown in supplementary materials \ref{file:DBschema}

\subsection{Feature extraction and feature engineering}\label{FeatureEngineering}
The following features were extracted from the database: (1) demographic data: gender, first visit age; (2) clinical data: whether the patient having concomitant inducible urticaria; (3) laboratory data: results from common blood tests, CRP, immunoglobulin, 25-hydroxyvitamin D, Thyroid function, autoantibodies, coagulation function, common urine tests, and allergen specific IgE tests.

For laboratory data, 2 types of data were extracted: time-independent data and time-dependent data. Time-independent data are average values of laboratory data during the whole follow-up period, while time-dependent data are average values of laboratory data durting preclincial phase (before the onset of urticaria), acute phase (within 6 weeks after the onset of urticaria), and chronic phase (after 6 weeks of the onset of urticaria).

Two datasets were compared for prediction performance in the model development process. The Sql queries for data are shown in in supplementary materials \ref{file:sql} and \ref{file:sql_time}. For each dataset, features with serious missing data (missing rate > 95\%) were excluded. Patients with over 50\% not null values with the left features were included in the dataset, and the missing values were imputed with the median value of the feature.

Visiting duration, calculated by the difference between the first visit date and the last visit date in the urticaria specialty clinic, was used as surrogate for remission time of urticaria, the validity of which will be analyzed in results section\ref{visit_duration}. Two binary classification threshold were set for visiting duration: 100 days and 365 days, based on natural history of chronic urticaria shown in previous studies\citep{eun2019natural,curto2018clinical,gaig2004epiodemiology}.

\subsection{Model development and comparison}\label{Training}
Dataset was split into training set and external test set with a ratio of 7:3. 5 models were adopted for comparison: Xgboost, random forest, gradient boosting machine (GBM), adaboost, and support vector machine (SVM). Time-dependent data extraction may introduce multicollinearity between features, which can affect model performance. To address this issue, L1 regularization, L2 regularization, and column sampling methods were employed in XGBoost to handle multicollinearity. L1 regularization helps eliminate irrelevant collinear features by sparse feature selection, while L2 regularization reduces the model's dependence on collinear features by smoothing the weights. Column sampling reduces the likelihood of collinear features being selected simultaneously by randomly choosing a subset of features when building each tree, which is helpful when interpreting the model using method like SHAP, since it enables more balanced weight distribution across features and reduces the possibility of one feature overshadowing others. The Random Forest model does not support L1 and L2 regularization, but it handles multicollinearity through the max\_features parameter, which is similar to column sampling. L2 regularization was applied to the SVM model as well. Additionally, the frameworks of XGBoost and GBM incorporate greedy split and boosting mechanisms, selecting the most important features for each tree and fitting the residuals step by step, allowing the model to focus on the key features rather than over-relying on collinear ones.

Hyperparameter optimization on 10 model settings (5 models and 2 datasets) was performed using TPE algorithm by nni package in python, which is a bayesian optimization algorithm that uses tree-structured parzen estimator to model the objective function and suggest the next set of hyperparameters to evaluate based on the previous results. Internal 5 fold cross-validation was employed to discern the most suitable hyperparameters for each distinct model, individually applied to each model settings for enhanced performance. The model's performance was assessed using ROC-AUC, accuracy, precision, recall and F1 score on different cuttoffs of visiting duration. The sensitivity, specificity, PPV, NPV, accuracy, and F1 score were calculated at the optimal cutoff value that maximized the Youden index. The best-performing model and data were selected for further analysis.

@@

In this study, Boruta algorithm was performed on the final model, with max iteration set to 50 and repeatition for 15 times. The confirmed features list and importance rank were recorded for each run. The overlap of confirmed features across runs was calculated to determine the final feature list. For time-dependent data, once a laboratory item was confirmed as important in any phase of disease, the feature of the same lab item in other phase of disease was also considered as important. That is to say, when top 10 items were selected, the features representing different phases of that item were all included in the final model, which is at most 30 features for time-dependent data. The model performance on different number of top items, ranked by the average importance score, were compared to determine the optimal number of features for the final model.

\subsection{Model explanation}\label{ModelExplanationMethods}
SHAP (Shapley Additive Explanations) is an explanation method based on game theory, used to interpret the outputs of machine learning models. It assigns each feature's contribution to the model's prediction, helping users understand the decision-making process of the model. In tree models that are not sensative to collinearity, such as Xgboost and GBM, the model tend to select one most important feature among collinear features for a split and ignores the other correlated features. As a result, SHAP may assign a high contribution to the first selected feature while the others receive little or no contribution. This happens because the model captures most of the information through the first feature, effectively overshadowing the others. By introducing joint distribution estimation for correlated features during SHAP calculations, we can capture their dependencies and ensure that SHAP values reflect the true impact of correlated features \citep{lundberg2017unified}.
The shapley value of each laboratory item in different phase of disease were compared to reveal various predicting ability. Some of the features were further analyzed with different age populations to explore the potential age effect on the predicting ability of laboratory data. 
Kernal density estimation was used to visualize the distribution of laboratory data in different phase of disease to provide a reference for the interpretation of SHAP values.

\section{Results}\label{Results}
\subsection{Visit duration as a validate surrogate for time to remission}\label{surrogate}

To verify visit duration as a surrogate for natural time to remission, we analyzed the distribution of visit duration in the dataset. The distribution of visit duration is shown in Figure \ref{visit_duration}. Similar to epidemiologic study which reported that half of the patients with chronic urticaria achieved remission after 11 months, and approximately 90\% achieved remission within 5 years\citep{eun2019natural}, the distribution of visit duration in our dataset showed that approximately 68\% of all patients have a visit duration more than 6w. Within those patients, 54\% of them have a visit duration less then a year, and rest of them have a visit duration more than a year. Unfortunately, our dataset does not include any patients with a visit duration exceeding five years. This study's visit duration distribution reflects the chronic urticaria remission pattern, suggesting that visit duration serves as a reliable proxy for remission time.

% insert image
\begin{figure}[t]
    \centering
    \includegraphics[width=0.8\linewidth]{figures/outcome_distribution.png}
    \caption{Distribution of visit duration in the dataset. The x-axis represents the visit duration in days, and the y-axis represents the number of patients. lines with different colors represent different cutoffs of visit duration. The percentage of patients within different cutoffs are shown in the legend.}\label{visit_duration}
\end{figure}

@@

Of 9921 patients recruited for this study, 3954 patients were selected for time-independent dataset and 1948 patients were selected for time-dependent dataset, the rest were excluded due to serious missing data as describes in methods section. Two datasets were allocated into seperate derivation and external validation sets with a ratio of 7:3, respectively. The comparison of demographic and clinical variables among the training and external validation sets is shown in Table \ref{tab:train_test_origi} for time-independent dataset and Table \ref{tab:train_test_time} for time-dependent dataset. No substantial differences were observed between the training set and the external test set across the majority of features in either time-independent or time-dependent data. Details of the study design are displayed in Figure \ref{protocol}.

% insert image
\begin{figure}[t]
    \centering
    \includegraphics[width=0.8\linewidth]{figures/protocol.png}
    \caption{Study design}\label{protocol}
\end{figure}

The comparison of clinical characteristics between patients with short and prolonged visit duration in the time-independent dataset (Table \ref{tab:good_outcome_poor_outcome_origi}) highlights several trends that align with findings from previous studies on chronic spontaneous urticaria (CSU).
Patients with visiting duration more than 100 days were defined as poor outcome group, and patients with visiting duration less than 100 days were defined as good outcome group. In this analysis, immunoglobulin E (IgE) ($131.65 \pm 242.00$ in good outcome vs $152.28 \pm 346.75$ in poor outcome), and basophil percentage ($0.25 \pm 0.22$ in good outcome vs $0.32 \pm 0.24$ in poor outcome) were significantly higher in patients with poor outcomes, which is consistent with previous studies that have linked these markers to CSU severity and duration \citep{SanchezBorges2017Factors,Rabelo-Filardi2013Parameters}.

In contrast, C-reactive protein (CRP) levels were significantly lower in patients with poor outcomes ($13.03 \pm 19.62$ in good outcome vs $7.71 \pm 10.23$ in poor outcome), which contradicts previous findings that have associated higher CRP levels with CSU severity \citep{Rabelo-Filardi2013Parameters}. However, this discrepancy could reflect the complexity of CRP as an inflammatory marker that may vary across different disease stages or subgroups of patients. In time-dependent data, which is shown in supplementary Table \ref{tab:good_outcome_poor_outcome_time}, CRP presented various trends in different phases of disease, with significantly lower CRP levels in acute phase ($10.21 \pm 10.95$ in good outcome vs $8.95 \pm 8.22$ in poor outcome) and higher CRP levels in chronic phase ($5.39 \pm 4.84$ in good outcome vs $6.47 \pm 8.81$ in poor outcome) in patients with poor outcomes. This finding suggests that the predictive ability of CRP may be influenced by the disease phase, which could explain the inconsistent results observed in the time-independent dataset.

@@

    \makecell[l]{C-reactive Protein} & $13.03 \pm 19.62$ & $7.71 \pm 10.23$ & 0.000 $\downarrow$ \\

@@

    \makecell[l]{Ro 52} & $2.10 \pm 5.90$ & $2.23 \pm 6.81$ & 0.515  \\
    \hline\end{tabular}\caption{Comparison of the characteristics between patients with good and poor outcomes in the time-independent dataset \\ continuous variables are presented as mean ± standard deviation, categorical variables are presented as number (percentage) \\ good outcome is defined as visit duration < 100 days, poor outcome is defined as visit duration $\geq$ 100 days} \label{tab:good_outcome_poor_outcome_origi}
    \end{table}

\subsection{Comparison of multiple models on time-dependent and time-independent data}\label{ModelComparison}
% xgboost + time-dependent datasset is the best model in 42 100 365 cutoffs

Time-independent and Time-dependent data, combined with 5 different models (XGBoost, Random Forest, AdaBoost, GBM, SVM), were used to generate 10 machine learning models to predict the visiting duration of urticaria. Among the 10 models, the XGBoost model with time-dependent data (AUC = $0.932 \pm 0.016$ for 100, AUC = $0.860 \pm 0.028$ for 365) has the best predictive effect for visiting duration, followed by the Random Forest model with time-dependent data (AUC = $0.922 \pm 0.028$ for 100, AUC = $0.837 \pm 0.037$ for 365) as shown in Figure \ref{kfoldfirst3}.

Notably, time-dependent data consistently outperformed time-independent data across all models, indicating that the phase-specific laboratory data are more informative for predicting the visiting duration of urticaria.

The discriminative performances and ROC curves of all 10 models are listed in Supplementary Table \ref{tab:kfold_results}, and Supplementary Figure \ref{fig:kfold_all}, respectively.

% insert image
\begin{figure}[t] 
    \centering 
    \includegraphics[width=0.8\linewidth]{figures/kfoldfirst3.png} 
    \caption{ROC curves for 5-fold cross-validation results for three different models (XGBoost, Random Forest, and GBM) at 2 binary classification thresholds (100 and 365) in panel A and B. The models' performances are plotted with time-independent and time-dependent datasets, where red indicates the time-independent dataset and blue represents the time-dependent dataset. The shaded areas represent the 95\% confidence intervals.}\label{kfoldfirst3} 
\end{figure}

@@

Xgboost with time-dependent data, as the best performing model, was selected for further optimization. The Boruta algorithm identified 38 laboratory items as important, as shown in Supplementary Table \ref{tab:boruta_confirmed_vars}. Feature importance ranking by Boruta algorithm are shown in Figure \ref{boruta_by_group} and Supplementary Table \ref{tab:boruta_ranking_df}. We rank the confirmed features by the average importance score and compare the model performance on different number of top items.

During the process of reducing features based on the feature importance, the changes in AUCs for the model shows that the model maintain a robust performance unitil the number of items reduced to less than 25, as shown in Figure \ref{topn_extval25} A.

The 25-item model were selected as the final model, and showed a stable performance in external test set with AUC of 0.956 and 0.877 for 100 and 365 cutoffs, as shown in Figure 3 \ref{topn_extval25} B.

% insert image
\begin{figure}[t] 
    \centering
    \includegraphics[width=0.8\linewidth]{figures/boruta_by_group.png} 
    \caption{Boxplot representing averaged feature ranking distribution for each laboratory item generated by the Boruta algorithm over 15 iterations. The ranking value for each laboratory item is averaged for 3 different phases of disease. The y-axis shows the importance rank, with lower values indicating higher importance, while the x-axis lists the laboratory items names.}\label{boruta_by_group}
\end{figure}

% insert image
\begin{figure}[t]  
    \centering 
    \includegraphics[width=0.8\linewidth]{figures/extvaltopn.png} 
    \caption{Panels A display the ROC curves for the model as the number of top items is reduced based on the average importance score at cutoff values of 100 and 365, respectively. Panels B show the ROC curves for the final model on external test set at cutoff values of 100 and 365, respectively.}\label{topn_extval25}

@@

\subsection{Model explanation}\label{ModelExplanationResults}
SHAP analysis was conducted on the final model to evaluate the feature importance. Figure \ref{heatmap_all_orderx} presents a heatmap of SHAP values from 100 randomly selected samples, ordered by hierarchical clustering to group similar patterns. The heatmap illustrates how these features influence the model’s output, with global feature importance displayed as a bar plot.

\begin{figure}[t] 
    \centering
    \includegraphics[width=0.8\linewidth]{figures/heatmap_all_orderx.png}
    \caption{a heatmap plot of SHAP values of all features on 100 randomly selected instance, with the instances on the x-axis, the model inputs on the y-axis, and the SHAP values encoded on a color scale. The samples are ordered based on a hierarchical clustering by their explanation similarity. This results in samples that have the same model output for the same reason getting grouped together. The output of the model is shown above the heatmap matrix, and the global importance of each model input shown as a bar plot on the right hand side of the plot.}\label{heatmap_all_orderx}
\end{figure}

@@

At last, autoimmunity markers, such as Anti Jo 1 and Ro 52, are having positive effect on the model output, as shown in Figure \ref{modelexplanation} I and J.

Other features that were not mentioned in the previous studies, such as mean corpuscular hemoglobin concentration, mean corpuscular volume, and platelet distribution width, also showed significant effects on the model output, as shown in Supplementary Figure \ref{modelexplanation2}.

KDE plots of features mentioned above in different phases of disease are shown in Supplementary Figure \ref{modelexplanation_kde} to illustrate the feature effect from the data perspective.

% insert image
\begin{figure}[t] 
    \centering
    \includegraphics[width=0.8\linewidth]{figures/modelexplanation.png} 
    \caption{SHAP value summary plots for well-known laboratory markers across different disease phases (preclinical, acute, and chronic) in all age groups. Each dot represents a patient sample, with the x-axis indicating the SHAP value (the feature’s impact on the model output) and the color gradient representing the feature value (red for high values, blue for low values). Features with higher SHAP values indicate a greater positive impact on predicting longer visit durations, while negative SHAP values correspond to shorter visit durations.}\label{modelexplanation}
\end{figure}

@@

\href{run:data.sql}{Supplementary file 2: SQL queries for time-independent extraction}
\label{file:sql}

\href{run:data_time.sql}{Supplementary file 3: SQL queries for time-dependent extraction}
\label{file:sql_time}
